\section{Architecture}

\subsection{High-Level Design}

The system consists of four components arranged in a sequential pipeline:

\begin{enumerate}[noitemsep]
    \item \textbf{Data pipeline:} Fetches historical lap data via FastF1, applies compound and track status filters, engineers features, and normalizes per race.
    \item \textbf{Sequence builder:} Constructs sliding-window input sequences (10 laps $\times$ 9 features) and delta lap time targets per driver stint.
    \item \textbf{GRU model (per compound):} Three independent GRU networks trained on sequences filtered by tire compound (SOFT, MEDIUM, HARD).
    \item \textbf{Strategy simulator:} At inference time, loads a 2024 race, generates worn and fresh tire predictions for each candidate pit lap, and identifies the first lap where pitting is net positive.
\end{enumerate}

\subsection{Data Pipeline}

\subsubsection{Data Source and Filtering}

Race data is fetched using FastF1 v3.8.2 for the 2022 (22 rounds), 2023 (22 rounds), and 2024 (24 rounds) seasons. The 2024 season is held out entirely for evaluation; models are trained exclusively on 2022--2023 data.

Three filters are applied to each race:
\begin{enumerate}[noitemsep]
    \item \texttt{IsAccurate == True}: removes laps with timing errors
    \item \texttt{TrackStatus == '1'}: retains only green-flag laps (removes safety car, yellow flag, and red flag laps)
    \item \texttt{PitInTime.isna()}: removes the pit-entry lap where the driver lifts off
\end{enumerate}

Only dry-compound laps (SOFT, MEDIUM, HARD) are retained; intermediate and wet tires are excluded. Weather data (air temperature, track temperature) is merged onto each lap via \texttt{pd.merge\_asof} on lap start time.

\subsubsection{Feature Engineering}

The input feature vector for each lap consists of 9 features:

\begin{table}[h]
\centering
\caption{Input Features}
\begin{tabular}{lll}
\toprule
Feature & Description & Normalization \\
\midrule
LapTimeSeconds & Lap time (s) & Per-race MinMax \\
StintLength & Laps on current tyre set & Per-race MinMax \\
FuelLoad & Estimated fuel remaining (kg) & Per-race MinMax \\
AirTemp & Ambient air temperature (\textdegree C) & Per-race MinMax \\
TrackTemp & Track surface temperature (\textdegree C) & Per-race MinMax \\
Compound\_SOFT & 1 if on SOFT tyres & Binary \\
Compound\_MEDIUM & 1 if on MEDIUM tyres & Binary \\
Compound\_HARD & 1 if on HARD tyres & Binary \\
TrackTemp\_global & TrackTemp scaled across all training races & Global MinMax \\
\bottomrule
\end{tabular}
\end{table}

Fuel load is estimated as $\max(0, 110 - \text{LapNumber} \times 1.6)$ kg, using a standard burn rate of 1.6 kg/lap from a 110 kg fuel load. Per-race MinMax normalization is applied to continuous features using scikit-learn's \texttt{MinMaxScaler} to account for the wide variation in baseline lap times across circuits (e.g., Monaco 75 s vs.\ Bahrain 97 s). The \texttt{TrackTemp\_global} feature preserves the absolute temperature signal across circuits after per-race normalization removes it from the standard TrackTemp column.

\subsubsection{Sequence Building}

For each driver stint, a sliding window of width 10 generates training samples. A window starting at lap $i$ produces input $X_i \in \mathbb{R}^{10 \times 9}$ and target $y_i = t_{i+10} - t_{i+9}$ (the lap-to-lap delta of the next lap). Stints shorter than 10 laps produce no samples and are discarded. After splitting by compound, the training dataset contains:

\begin{table}[h]
\centering
\caption{Training Sequences by Compound (2022--2023)}
\begin{tabular}{lrr}
\toprule
Compound & Train Sequences & Test Sequences (2024) \\
\midrule
SOFT   & 4,087  & 200 \\
MEDIUM & 9,853  & 442 \\
HARD   & 15,041 & 2,584 \\
\midrule
Total  & 28,981 & 3,226 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Model Architecture}

Two recurrent architectures were evaluated: TireLSTM and TireGRU. Both take a sequence of shape $(B, 10, 9)$ as input and produce a scalar delta prediction.

\textbf{TireGRU} (selected final architecture):
\begin{align}
\mathbf{h}_t &= \text{GRU}(\mathbf{x}_t, \mathbf{h}_{t-1}) \quad \in \mathbb{R}^{64} \\
\hat{y} &= \text{Linear}(\text{Dropout}(\mathbf{h}_{10})) \quad \in \mathbb{R}
\end{align}

Configuration: hidden size 64, 2 layers, dropout 0.2, batch-first. Total parameters: approximately 50,000. Training uses Adam optimizer \cite{kingma2015} with learning rate $10^{-3}$, ReduceLROnPlateau scheduler (patience=3, factor=0.5), gradient clipping at norm 1.0, and early stopping at patience=7 on validation loss.

An LSTM with additive attention \cite{bahdanau2015} was also evaluated. Attention computes a learned weighted sum over all 10 hidden states before the output projection. This variant was the worst performer (MAE $-22.5\%$ vs.\ baseline), as discussed in Section 4.3.

\subsection{Strategy Simulator}

At inference time, the simulator loads a 2024 race for a specific driver, normalizes the data using the same per-race scaler, and evaluates each candidate pit lap as follows:

\textbf{Worn tire projection} (\texttt{predict\_worn}): Starting from the last 10 observed laps, the model auto-regressively predicts 20 future laps. StintLength increments by 1 at each step; FuelLoad decrements proportionally; compound one-hot remains constant.

\textbf{Fresh tire projection} (\texttt{predict\_fresh}): Seeds from the earliest low-tyre-life laps of the target compound (StintLength $< 3$ normalized laps) from earlier in the race, resetting StintLength to 0.

\textbf{Decision criterion}: For each candidate pit lap $\ell$, compute:
\begin{equation}
\Delta(\ell) = \underbrace{L_\text{pit} + \sum_{k=1}^{20} \hat{t}_k^\text{fresh}}_{\text{pit cost}} - \underbrace{\sum_{k=1}^{20} \hat{t}_k^\text{worn}}_{\text{stay cost}}
\end{equation}
where $L_\text{pit}$ is the normalized pit stop time loss (22 seconds / max lap time; circuit-specific values for Monaco, Singapore, Las Vegas, and Miami). The recommended pit lap is the first $\ell$ where $\Delta(\ell) < 0$. One-stop and two-stop strategies are evaluated by brute-force search over all valid $(p_1)$ and $(p_1, p_2)$ pit lap pairs; the total race time under each strategy determines which is preferred.
