\begin{thebibliography}{99}

\bibitem{fastf1}
T.\ Pretzel, ``FastF1: A Python package for accessing Formula 1 data,'' GitHub repository, 2024. [Online]. Available: \url{https://github.com/theOehrly/Fast-F1}

\bibitem{hochreiter1997}
S.\ Hochreiter and J.\ Schmidhuber, ``Long short-term memory,'' \textit{Neural Computation}, vol.\ 9, no.\ 8, pp.\ 1735--1780, 1997.

\bibitem{cho2014}
K.\ Cho, B.\ van Merrienboer, C.\ Gulcehre, D.\ Bahdanau, F.\ Bougares, H.\ Schwenk, and Y.\ Bengio, ``Learning phrase representations using RNN encoder-decoder for statistical machine translation,'' in \textit{Proc.\ EMNLP}, 2014, pp.\ 1724--1734.

\bibitem{chung2014}
J.\ Chung, C.\ Gulcehre, K.\ Cho, and Y.\ Bengio, ``Empirical evaluation of gated recurrent neural networks on sequence modeling,'' in \textit{NIPS Workshop on Deep Learning}, 2014.

\bibitem{bahdanau2015}
D.\ Bahdanau, K.\ Cho, and Y.\ Bengio, ``Neural machine translation by jointly learning to align and translate,'' in \textit{Proc.\ ICLR}, 2015.

\bibitem{kingma2015}
D.\ P.\ Kingma and J.\ Ba, ``Adam: A method for stochastic optimization,'' in \textit{Proc.\ ICLR}, 2015.

\bibitem{heilmeier2020}
A.\ Heilmeier, W.\ Wischnewski, L.\ Hermansdorfer, J.\ Betz, M.\ Lienkamp, and B.\ Lohmann, ``Minimum curvature trajectory planning and control for an autonomous race car,'' \textit{Vehicle System Dynamics}, vol.\ 58, no.\ 10, pp.\ 1497--1527, 2020.

\bibitem{mcaleer2023}
S.\ McAleer, A.\ Lamont, S.\ Deans, and R.\ Savani, ``Approximate exploitability: Learning a best response in large extensive-form games,'' in \textit{Proc.\ NeurIPS}, 2023.

\bibitem{tulabandhula2014}
T.\ Tulabandhula and C.\ Rudin, ``On combining machine learning with decision making,'' \textit{Machine Learning}, vol.\ 97, no.\ 1, pp.\ 33--64, 2014.

\bibitem{cheng2021}
Y.\ Cheng and M.\ Savani, ``Learning to race in extreme heat: Reinforcement learning for endurance racing strategy,'' in \textit{Proc.\ IJCAI Workshop on AI in Sport}, 2021.

\bibitem{loiacono2011}
D.\ Loiacono, P.\ L.\ Lanzi, J.\ Togelius, E.\ Onieva, D.\ A.\ Pelta, M.\ V.\ Butz, T.\ D.\ L{\"{o}}nneker, L.\ Cardamone, D.\ Pérez, Y.\ S{\'{a}}ez, M.\ Preuss, and J.\ Quadflieg, ``The 2009 simulated car racing championship,'' \textit{IEEE Transactions on Computational Intelligence and AI in Games}, vol.\ 2, no.\ 2, pp.\ 131--147, 2010.

\bibitem{vaswani2017}
A.\ Vaswani, N.\ Shazeer, N.\ Parmar, J.\ Uszkoreit, L.\ Jones, A.\ N.\ Gomez, L.\ Kaiser, and I.\ Polosukhin, ``Attention is all you need,'' in \textit{Proc.\ NeurIPS}, 2017.

\end{thebibliography}
